Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-04-12 doc/mc-paper.tex
More consistent notation

The process for generating such a proof depends on the form of node $\nodeu$:
\begin{enumerate}
\item If $\nodeu$ is a literal $\lit'$, then the formula
  $\simplify{\psi}{\passign}$ must consist of the single unit clause
  $C = \{\lit'\}$, such that any $C' \in \prov_{\passign}(C, \psi)$ must have $C' \subseteq \{ \lit' \} \cup\, \bigcup_{\lit \in \passign} \obar{\lit}$.
  Any of these can
  serve as the target clause.
\item If $\nodeu$ is a sum node with children $\nodeu_1$ and $\nodeu_0$,
  then, since the node originated from a d-DNNF graph, there must be
  some variable $x$ such that either $\nodeu_1$ is a literal node for $x$ or $\nodeu_1$ is a
  product node containing a literal node for $x$ as a child.  In either case, we
  recursively call $\validate(\nodeu_1, \passign \cup \{ x \}, \psi)$.
  This will cause the addition of the target clause
  $u_1 \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
Similarly, either $\nodeu_0$ is a literal node for $\obar{x}$ or $\nodeu_0$ is a product node containing a literal node for $\obar{x}$ as
  a child.  In either case, we recursively call $\validate(\nodeu_0, \passign \cup \{ \obar{x} \}, \psi)$,
  causing the addition of the target clause
  $u_0 \lor x \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
  These recursive results can be combined with the second and third defining clauses for $\nodeu$
(see Table~\ref{tab:defining}(B))
  to generate the target clause for $\nodeu$.
\item If $\nodeu$ is a product node, then we can divide its children
  into a set of literal nodes $L$ and a set of nonterminal nodes $\nodeu_1, \nodeu_2, \ldots, \nodeu_k$.
  \begin{enumerate}
    \item For each literal
  $\lit \in L$, we must prove that any total assignment $\alpha$, such that
  $\passign \subset \alpha$ has $\alpha(\lit) = 1$.  In some
  cases, this can be done by simple Boolean constraint propagation (BCP).
  In other cases, we must prove that the formula
  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ is unsatisfiable.  We
  do so by writing the formula to a file, invoking a proof-generating
  SAT solver, and then converting the generated unsatisfiability proof
  into a sequence of clause additions in the CPOG file.
\item For a single nonterminal child ($k = 1$), we recursively call
  $\validate \left(\nodeu_1, \passign \cup L, \psi\right)$.
\item For multiple nonterminal children ($k > 1$),
  it must be the case that the clauses in
  $\gamma = \simplify{\psi}{\passign}$ can be partitioned into $k$ subsets
  $\gamma_1, \gamma_2, \ldots, \gamma_k$ such that $\dependencyset(\gamma_i)
  \cap \dependencyset(\gamma_j) = \emptyset$ for $1 \leq i < j \leq k$,
  and we can match each node $\nodeu_i$ to subset $\gamma_i$ based on its
  literals.
  For each $i$ such that $1 \leq i \leq k$, let $\psi_i = \prov_{\passign}(\gamma_i, \psi)$, i.e., those input clauses in $\psi$ that, when simplified, became clause partition $\gamma_i$.
  We recursively call
  $\validate \left(\nodeu_i, \passign \cup L, \psi_i\right)$.
\end{enumerate}
  We then generate the target clause for node $\nodeu$,
creating the hint by combining the results from the BCP and SAT calls for
  the literals, the recursively computed target clauses, and all but
  the first defining clause for node $\nodeu$
(see Table~\ref{tab:defining}(A)).
\end{enumerate}
Most of these steps involve a polynomial number of
operations per recursive call, with the exception of those that call
a SAT solver to validate a literal.

@@

{\em Literal Grouping:} A single recursive step of $\validate$ can encounter product nodes
having many---tens or even hundreds---of literals as children.  The
earlier formulation of $\validate$ considers each literal $\lit \in L$
separately, calling a SAT solver for every literal that cannot be
validated with BCP\@.  Literal grouping handles
all of these literals together.  It defines product node $\nodev$
having the literals as children.  The goal then becomes to prove that
any total assignment must yield 1 for extension variable $v$.  Calling
a solver with $v$ set to 0 yields an unsatisfiability proof that can
be mapped back to a sequence of clause additions in the CPOG file validating
all of the literals.

@@

A single recursive step of $\validate$ can encounter product nodes
having many literals as children.  The naive formulation of $\validate$
considers each literal $\lit \in L$ separately.
Literal grouping allows all literals to be validated with a single call to a SAT solver.
It collects those literals
$\lit_1, \lit_2, \ldots, \lit_m$ that cannot be validated by BCP and defines a
product node $\nodev$ having these literals as children.  The goal
then becomes to prove that any total assignment must yield 1 for extension
variable $v$.  A single call to the solver can generate this proof by invoking it on the formula
  $\simplify{\psi}{\passign} \cup \simplify{\theta_{v}}{\{ \obar{v} \}}$, which should be unsatisfiable.
  The proof steps can be mapped back into clause addition steps in the CPOG file, incorporating the
  input clauses and the defining clauses for $\nodev$ into the hints.

>>>>

The process for generating such a proof depends on the form of node $\nodeu$:
\begin{enumerate}
\item If $\nodeu$ is a literal $\lit'$, then the formula
  $\simplify{\psi}{\passign}$ must consist of the single unit clause
  $C = \{\lit'\}$, such that any $C' \in \prov_{\passign}(C, \psi)$ must have $C' \subseteq \{ \lit' \} \cup\, \bigcup_{\lit \in \passign} \obar{\lit}$.
  Any of these can
  serve as the target clause.
\item If $\nodeu$ is a sum node with children $\nodeu_1$ and $\nodeu_0$,
  then, since the node originated from a d-DNNF graph, there must be
  some variable $x$ such that either $\nodeu_1$ is a literal node for $x$ or $\nodeu_1$ is a
  product node containing a literal node for $x$ as a child.  In either case, we
  recursively call $\validate(\nodeu_1, \passign \cup \{ x \}, \psi)$.
  This will cause the addition of the target clause
  $u_1 \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
Similarly, either $\nodeu_0$ is a literal node for $\obar{x}$ or $\nodeu_0$ is a product node containing a literal node for $\obar{x}$ as
  a child.  In either case, we recursively call $\validate(\nodeu_0, \passign \cup \{ \obar{x} \}, \psi)$,
  causing the addition of the target clause
  $u_0 \lor x \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
  These recursive results can be combined with the second and third defining clauses for $\nodeu$
(see Table~\ref{tab:defining}(B))
  to generate the target clause for $\nodeu$.
\item If $\nodeu$ is a product node, then we can divide its children
  into a set of literal nodes $\lambda$ and a set of nonterminal nodes $\nodeu_1, \nodeu_2, \ldots, \nodeu_k$.
  \begin{enumerate}
    \item For each literal
  $\lit \in \lambda$, we must prove that any total assignment $\alpha$, such that
  $\passign \subset \alpha$ has $\alpha(\lit) = 1$.  In some
  cases, this can be done by simple Boolean constraint propagation (BCP).
  In other cases, we must prove that the formula
  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ is unsatisfiable.  We
  do so by writing the formula to a file, invoking a proof-generating
  SAT solver, and then converting the generated unsatisfiability proof
  into a sequence of clause additions in the CPOG file.
\item For a single nonterminal child ($k = 1$), we recursively call
  $\validate \left(\nodeu_1, \passign \cup \lambda, \psi\right)$.
\item For multiple nonterminal children ($k > 1$),
  it must be the case that the clauses in
  $\gamma = \simplify{\psi}{\passign}$ can be partitioned into $k$ subsets
  $\gamma_1, \gamma_2, \ldots, \gamma_k$ such that $\dependencyset(\gamma_i)
  \cap \dependencyset(\gamma_j) = \emptyset$ for $1 \leq i < j \leq k$,
  and we can match each node $\nodeu_i$ to subset $\gamma_i$ based on its
  literals.
  For each $i$ such that $1 \leq i \leq k$, let $\psi_i = \prov_{\passign}(\gamma_i, \psi)$, i.e., those input clauses in $\psi$ that, when simplified, became clause partition $\gamma_i$.
  We recursively call
  $\validate \left(\nodeu_i, \passign \cup \lambda, \psi_i\right)$.
\end{enumerate}
  We then generate the target clause for node $\nodeu$,
creating the hint by combining the results from the BCP and SAT calls for
  the literals, the recursively computed target clauses, and all but
  the first defining clause for node $\nodeu$
(see Table~\ref{tab:defining}(A)).
\end{enumerate}
Most of these steps involve a polynomial number of
operations per recursive call, with the exception of those that call
a SAT solver to validate a literal.

@@

{\em Literal Grouping:} A single recursive step of $\validate$ can encounter product nodes
having many---tens or even hundreds---of literals as children.  The
earlier formulation of $\validate$ considers each literal $\lit \in \lambda$
separately, calling a SAT solver for every literal that cannot be
validated with BCP\@.  Literal grouping handles
all of these literals together.  It defines product node $\nodev$
having the literals as children.  The goal then becomes to prove that
any total assignment must yield 1 for extension variable $v$.  Calling
a solver with $v$ set to 0 yields an unsatisfiability proof that can
be mapped back to a sequence of clause additions in the CPOG file validating
all of the literals.

@@

A single recursive step of $\validate$ can encounter product nodes
having many literals as children.  The naive formulation of $\validate$
considers each literal $\lit \in \lambda$ separately.
Literal grouping allows all literals to be validated with a single call to a SAT solver.
It collects those literals
$\lit_1, \lit_2, \ldots, \lit_m$ that cannot be validated by BCP and defines a
product node $\nodev$ having these literals as children.  The goal
then becomes to prove that any total assignment must yield 1 for extension
variable $v$.  A single call to the solver can generate this proof by invoking it on the formula
  $\simplify{\psi}{\passign} \cup \simplify{\theta_{v}}{\{ \obar{v} \}}$, which should be unsatisfiable.
  The proof steps can be mapped back into clause addition steps in the CPOG file, incorporating the
  input clauses and the defining clauses for $\nodev$ into the hints.
